How do changes in ex vary between countries?

### 2022-03-03 draft_paper.Rmd
fail to make decent progress with tables; do lots of linear regression models on decadal differences

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

@@

pacman::p_load(here, tidyverse, flextable) # using flextable instead of gt. It seems to be better 
# developed, more commonly used, and has more clearly labelled function names.

@@

Some insightful comment about what's going on.

@@

## Change in life expectancy in these nations 
Next, we compare the annual change in life expectancy at birth and at aged 65 years for the 6 nations from 1980 to the latest available year, by sex.

@@

change_in_ex_selected_countries
```
### Changes in life expectancy at birth

@@

@@

# Time trends 
Finally, we used time-trend analysis to explore the changes in life expectancy over the 4 decades: 1980s, 1990s, 2000s, and 2010s. This produces the average change for the decade, whether negative or positive, the standard error and its statistical significance for each country by sex.

@@

hmd_ex_selected_countries_with_synth %>% 
    filter(year >= 1979) %>% 
    group_by(code, x, sex) %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
    nest() %>% 
    mutate(
      rwd_model = map(data, ~lm(delta_ex ~ 1, data = . )),
      ar_model  = map(data, ~lm(delta_ex ~ lag(delta_ex), data = .))
    ) %>% 
    mutate(
      aic_rwd = map_dbl(rwd_model, AIC),
      aic_ar  = map_dbl(ar_model, AIC)
    ) %>% 
    mutate(
      diff_aic = aic_ar - aic_rwd
    )  %>% 
  # filter(diff_aic > 0)
  mutate(
    tidied_ar_model = map(ar_model, broom::tidy)
  ) %>% 
  unnest(tidied_ar_model) %>% 
  filter(term == "(Intercept)") %>% 
  arrange(desc(estimate)) %>% 
  select(code, x, sex, estimate, std.error) %>% 
  mutate(display_label = glue::glue("{code}_{x}_{sex}")) %>% 
  ggplot(aes(estimate, fct_reorder(display_label, estimate))) + 
  geom_point()

Look at improvements by decade

hmd_ex_selected_countries_with_synth %>% 
    filter(year >= 1979) %>% 
    group_by(code, x, sex) %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>%
    mutate(decade = year - (year %% 10)) %>% 
    filter(!is.na(delta_ex)) %>% 
  group_by(code, x, sex, decade) %>% 
  summarise(
    mean_delta_ex = mean(delta_ex),
    sd_delta_ex   = sd(delta_ex)
    ) %>% 
  ungroup() %>% 
  mutate(cell_contents = glue::glue("{round(mean_delta_ex, 3)} ({round(sd_delta_ex, 3)})")
  ) %>% 
  filter(code == "GBRTENW") %>% 
  filter(x == 0) %>% 
  select(sex, decade, cell_contents) %>% 
  spread(decade, cell_contents) %>% 
  flextable()

>>>>

# To dos

- [ ] Create flextable by decade

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  tab.id = NULL,
  tab.cap = NULL,
  tab.topcaption = TRUE,
  tab.lp = "tab:"

  )
```

@@

pacman::p_load(knitr, here, tidyverse, flextable) # using flextable instead of gt. It seems to be better 
# developed, more commonly used, and has more clearly labelled function names.

@@

- [ ] Some insightful comment about what's going on.

@@

## Change in life expectancy in these nations

Next, we compare the annual change in life expectancy at birth and at aged 65 years for the 6 nations from 1980 to the latest available year, by sex.

@@

change_in_ex_selected_countries
```

### Changes in life expectancy at birth

- [ ] make smoother lines coloured by sex
- [ ] use consistent colours for all graphs showing separate sex results

@@

- [ ] Use slightly stronger (and consistent) colours for sex smoother lines

@@

## Time trends

Finally, we used time-trend analysis to explore the changes in life expectancy over the 4 decades: 1980s, 1990s, 2000s, and 2010s. This produces the average change for the decade, whether negative or positive, the standard error and its statistical significance for each country by sex.

- [X] Give this a title and further info (lower priority)
- [X] Look at splitting by starting age 
    - [ ] Look at further refining starting age split

```{r, echo = FALSE, warning = FALSE, message = FALSE}

hmd_ex_selected_countries_with_synth %>% 
    filter(year >= 1979) %>% 
    filter(!(code %in% c("DEUTE", "DEUTW", "DEUTNP"))) %>%  # Using only synthetic germany for longer time period
    group_by(code, x, sex) %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
    nest() %>% 
    mutate(
      rwd_model = map(data, ~lm(delta_ex ~ 1, data = . )),
      ar_model  = map(data, ~lm(delta_ex ~ lag(delta_ex), data = .))
    ) %>% 
    mutate(
      aic_rwd = map_dbl(rwd_model, AIC),
      aic_ar  = map_dbl(ar_model, AIC)
    ) %>% 
    mutate(
      diff_aic = aic_ar - aic_rwd
    )  %>% 
  # filter(diff_aic > 0)
  mutate(
    tidied_ar_model = map(ar_model, broom::tidy)
  ) %>% 
  unnest(tidied_ar_model) %>% 
  filter(term == "(Intercept)") %>% 
  arrange(desc(estimate)) %>%
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
  select(code, country, x, sex, estimate, std.error) %>% 
  mutate(display_label = glue::glue("{country}_{x}_{sex}")) %>% 
  ggplot(aes(estimate, fct_reorder(display_label, estimate))) + 
  geom_point() +
  expand_limits( x = 0) + 
  geom_vline(xintercept = 0) + 
  labs(
    x = "Average annual life expectancy change",
    y = "Country-sex combination",
    title = "Average annual improvements by country and sex",
    subtitle = "Range: 1980 to 2020 or nearest available years",
    caption = "Source: Human Mortality Database"
  ) + 
  facet_wrap(~x)

### Average improvements by decade

- [ ] Add further introduction 
- [ ] Add statistical significance (though maybe this is a different analysis)

```{r, tab.cap = "Source: HMD"}

hmd_ex_selected_countries_with_synth %>% 
    filter(year >= 1979) %>% 
    group_by(code, x, sex) %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>%
    mutate(decade = year - (year %% 10)) %>% 
    filter(!is.na(delta_ex)) %>% 
  group_by(code, x, sex, decade) %>% 
  summarise(
    mean_delta_ex = mean(delta_ex),
    sd_delta_ex   = sd(delta_ex)
    ) %>% 
  ungroup() %>% 
  mutate(cell_contents = glue::glue("{round(mean_delta_ex, 3)} ({round(sd_delta_ex, 3)})")
  ) %>% 
  filter(code == "GBRTENW") %>% 
  filter(x == 0) %>% 
  select(sex, decade, cell_contents) %>% 
  spread(decade, cell_contents) %>% 
#  write.csv("clipboard")
  flextable() %>%
  # Want to make the header row bold
  vline(j = 1, part = "all") %>%
  set_caption(caption = "Average improvement in life expectancy by decade")

Are the differences by decade statistically significant?

@@

dta_for_decade_comparison <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "female") %>% 
      filter(code == "GBRTENW") %>% 
      filter(year >= 1979) %>% 
      # ggplot(aes(x = year, y = ex)) +
      #   geom_line()
      arrange(year) %>% 
      mutate(delta_ex = ex - lag(ex)) %>% 
    #   ggplot(aes(x = year, y = delta_ex)) + 
    # geom_point() + geom_line() +
    # geom_hline(yintercept = 0)
      mutate(
        decade = year - (year %% 10)
      ) %>% 
  filter(!is.na(delta_ex))

model_null <- 
  lm(delta_ex ~ 1, data = dta_for_decade_comparison)

model_alt <- 
  lm(delta_ex ~ factor(decade), data= dta_for_decade_comparison)

summary(model_null)

summary(model_alt)

anova(model_null, model_alt)

# to get the value of interest do tmp[2,6]

# Generalising

tidy_dataset <- function(df){
  df %>% 
      filter(year >= 1979) %>% 
      arrange(year) %>% 
      mutate(delta_ex = ex - lag(ex)) %>% 
      mutate(
        decade = year - (year %% 10)
      ) %>% 
  filter(!is.na(delta_ex))
}

make_null_model <- function(df){
  lm(delta_ex ~ 1, data = df)
}

all_null_alt_models <-
  hmd_ex_selected_countries_with_synth %>%
    group_by(x, sex, code) %>% 
    nest() %>% 
    mutate(
      prep_data = map(data, tidy_dataset) 
    ) %>% 
    mutate(
      model_null = map(prep_data, ~lm(delta_ex ~ 1, data = .)),
      model_alt  = map(prep_data, ~lm(delta_ex ~ factor(decade), data = .))
  #    model_null = map(prep_data, make_null_model)
    ) %>% 
    mutate(
      anova_results = map2(model_null, model_alt, anova)
    ) %>% 
    mutate(
      anova_pval = map_dbl(anova_results, ~.[2,6])
    )

all_null_alt_models %>% 
  filter(anova_pval < 0.05) -> tmp

summary(tmp$model_alt[[1]])

summary(tmp$model_alt[[2]])

  # build null model
  # build alt model
  # do anova
  # get p value from anova
